---
title: "BBIO 495 Phagocytosis and Experimental Evolution Results"
author: "Elizabeth Ostrowski"
output:
  pdf_document: default
  html_notebook: default
---

You need to download from our Google Sheet the relevant worksheet as a .tsv (tab-separated values) data file.
You need to replace the filepath below with the filepath to this data file on your computer.

```{r}

# Install the tidyverse and other packages if you haven't done so previously
# install.packages('tidyverse') # run this one time to install tidyverse package, which enables use of ggplot functions

# Load libraries
library(tidyverse)
library(cowplot)
library(ggridges)
library(lme4)
library(emmeans)
library(lmerTest)

```


# Read in data
```{r}
# These are the collated data from our shared data entry pages, in long format
in_data_plq_size = read.table("~/Dropbox/Teaching/2025 Spring 495/Analyses/combined_plaque_size.tsv", sep="\t", header=T)
in_data_plq_num = read.table("~/Dropbox/Teaching/2025 Spring 495/Analyses/combined_plaque_number.tsv", sep="\t", header=T)
in_evol_spore_num = read.table("~/Dropbox/Teaching/2025 Spring 495/Analyses/evol_exp_total_spores.tsv", sep="\t", header=T)

# Update column names
names(in_data_plq_size) = c("group", "student_1", "student_2", "plate", "strain", "plaque_diam_pix", "plaque_diam_mm", "plaque_area", "exp_date", "image_date")

# Filter improbable values 
max_plaque = 50 # outliers, based on overall distribution
min_plaque = 1  # plaques smaller than 1 mm would not be visible in a photo; reflects a likely math error
plq_size_filt = in_data_plq_size %>% filter(plaque_diam_mm > min_plaque) %>% filter(plaque_diam_mm < max_plaque) %>% mutate(strain = fct_relevel(strain, "SBW25","Fuzzy", "Wrinkly", "Copycat", "Mucoid")) 

# TO DO: pair groups and remove those where SBW25 did not show any plaques

```


Plaque Diameter Histograms
```{r}
ggplot(data=plq_size_filt, aes(fill=strain)) + geom_histogram(aes(plaque_diam_mm)) + facet_wrap(vars(strain))

#ggplot(data=plq_size_filt, aes(x =plaque_diam_mm, y = strain, fill = stat(x))) +
#  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
#  scale_fill_viridis_c(name = "Plaque Diameter (mm)", option = "C") +
#  labs(title = 'Plaque Diameter (mm)') + xlab('"Plaque Diameter (mm)') + ylab('Strain')

ggplot(data=plq_size_filt, aes(x = plaque_diam_mm, y = strain, fill = strain)) +
  geom_density_ridges(aes(point_color = strain, point_fill = strain, point_shape = strain),
    alpha = 0.5, point_alpha = 1, jittered_points = TRUE) + 
  labs(title = 'Plaque Diameter (mm)') + xlab('"Plaque Diameter (mm)') + ylab('Strain')

#ggplot(data=plq_size_filt, aes(x = plaque_diam_mm, y = strain, fill = strain)) +
# geom_density_ridges(alpha=0.5) +
#  labs(title = 'Plaque Diameter (mm)') + xlab('"Plaque Diameter (mm)') + ylab('Strain')

```


Plaque Area Histograms (I think these aren't very helpful, since they are directly calculated from diameter)
```{r}
# Plaque Area
ggplot(data=plq_size_filt, aes(fill=strain)) + geom_histogram(aes(plaque_area)) + facet_wrap(vars(strain))

#ggplot(data=plq_size_filt, aes(x =plaque_area, y = strain, fill = stat(x))) +
#  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
#  scale_fill_viridis_c(name = "Plaque Area (mm^2)", option = "C") +
#  labs(title = 'Plaque Area (mm^2)') + xlab('"Plaque Area (mm^2)') + ylab('Strain')

#ggplot(data=plq_size_filt, aes(x = plaque_area, y = strain, fill = strain)) +
#  geom_density_ridges(alpha=0.5) +
#  labs(title = 'Plaque Area (mm^2)') + xlab('"Plaque Area (mm^2)') + ylab('Strain')

```

Summarise Plaque Size:
Average in steps: first the plaques on each plate, then average the two plates together
```{r}

psize_per_plate = plq_size_filt %>% group_by(group, strain, plate) %>% summarise(mean_diam = mean(plaque_diam_mm), mean_area = mean(plaque_area)) %>% mutate(strain = fct_relevel(strain, "SBW25","Fuzzy", "Mucoid", "Copycat", "Wrinkly"))

psize_per_group = psize_per_plate %>% group_by(group, strain) %>% summarise(mean_diam = mean(mean_diam), mean_area = mean(mean_area))

# All data (no averaging)
ggplot(data=plq_size_filt, aes(x=strain, y=plaque_area, col=strain)) + geom_boxplot(width=0.35,outlier.shape = NA) + geom_jitter(position=position_jitter(0.1),size=2, alpha=0.3)

# Averaged across plates
ggplot(data=psize_per_group, aes(x=strain, y=mean_diam, col=strain)) + geom_boxplot(width=0.35,outlier.shape = NA)+ geom_jitter(position=position_jitter(0.1),size=2) + ylab("Plaque Diameter (mm)")

# Averaged across plates
ggplot(data=psize_per_group, aes(x=strain, y=mean_area, col=strain)) + geom_boxplot(width=0.35,outlier.shape = NA)+ geom_jitter(position=position_jitter(0.1),size=2) + ylab("Plaque Area (mm^2)")
```

Statistics
```{r}
mod = lmer(plaque_diam_mm ~ strain + (1|group)*(1|exp_date), data=plq_size_filt, contrasts=T )
emmeans(mod, specs=trt.vs.ctrl ~ strain)

summary(mod)
anova(mod, test="Chisq") # Same conclusions as emmeans


```

Summarise Plaque Number
```{r}

in_data_plq_num$num_plaques = as.numeric(in_data_plq_num$num_plaques)
plq_num_filt = in_data_plq_num %>% mutate(strain = fct_relevel(strain, "SBW25","Wrinkly", "Copycat", "Mucoid", "Fuzzy"))

pnum_avg_across_plates = plq_num_filt %>% group_by(group, strain) %>% summarise(mean_plqs = mean(num_plaques, na.rm=T))

# Raw data
ggplot(data=plq_num_filt, aes(x=strain,y=num_plaques, col=strain)) + geom_boxplot(outlier.shape = NA) + geom_jitter(position=position_jitter(0.2),size=2) + ylab('Number of Plaques') + xlab("Strain")

# Averaged data
ggplot(data=pnum_avg_across_plates, aes(x=strain, y=mean_plqs, col=strain)) + geom_boxplot(width=0.3,outlier.shape = NA) + geom_jitter(aes(fill=strain), position=position_jitter(0.01),size=2) + ylab("Mean Number of Plaques") + xlab("Strain") 

```

Statistics

```{r}

all_data = pnum_avg_across_plates %>% left_join(psize_per_plate, by=c("group", "strain"))

ggplot(data=all_data, aes(x=mean_plqs, y=mean_area)) + geom_point() + xlab('Number of Plaques') + ylab('Plaque Area') + geom_smooth(method='lm') + ggtitle('Relationship Between Number of Plaques and their Size')

cor.test(all_data$mean_plqs, all_data$mean_area) # not significant


mod_pnum = lmer(num_plaques ~ strain + (1|group)*(1|exp_date), data=plq_num_filt, contrasts=T )
emmeans(mod_pnum, specs=trt.vs.ctrl ~ strain)

summary(mod_pnum)
anova(mod_pnum, test="Chisq")


```

Analyse Coevolution Experiment: Total Spores
```{r}

med_pop_size = in_evol_spore_num %>% group_by(round) %>% summarise(med_pop_size = median(total_spores))

# Automated y-axis
ggplot(data=in_evol_spore_num, col=factor(pop), group=pop) + geom_point(aes(x=round, y=total_spores,col=factor(pop), alpha=0.5)) + geom_line(aes(x=round, y=total_spores, col=factor(pop), alpha=0.5)) + ylab('Total Spores') + xlab('Evolution Round') + geom_point(data=med_pop_size, aes(x=round, y=med_pop_size), size=4)

# cutting off y-axis limit
ggplot(data=in_evol_spore_num, col=factor(pop), group=pop) + geom_point(aes(x=round, y=total_spores,col=factor(pop), alpha=0.5)) + geom_line(aes(x=round, y=total_spores, col=factor(pop), alpha=0.5)) + ylab('Total Spores') + xlab('Evolution Round') + geom_point(data=med_pop_size, aes(x=round, y=med_pop_size), size=4) + ylim(0,6e6)



```
Fix me: redo plots so that presentation reflects paired data (SBW25 + mutant per group)
```{r}

ggplot(pnum_avg_across_plates, aes(x = strain, y = mean_plqs, group = group)) +
  geom_line(alpha = 0.7, color = "gray") +
  geom_point(size = 2) +  theme_minimal() +
  labs(title = "Reaction Norm Plot", x = "Strain", y = "Plaque Number") +
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.x = element_blank())

ggplot(psize_per_group, aes(x = strain, y = mean_diam, group = group)) +
  geom_line(alpha = 0.7, color = "gray") +
  geom_point(size = 2) +  theme_minimal() +
  labs(title = "Reaction Norm Plot", x = "Strain", y = "Plaque Diammeter (mm)") +
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.x = element_blank())

```
